---
title: "Lab 5: Poisson Regression"
subtitle: "PSY 504: Advanced Statistics"
author: "Your Name"
date: "2026-03-18"
format: 
  html:
    code-fold: true
    code-overflow: wrap
    toc: true
    toc-depth: 2
    toc-location: left
execute: 
  message: false
  warning: false
engine: knitr
---

⬇️ [Download the .qmd](https://github.com/suyoghc/PSY504_Spring_2026/raw/main/posts/Poisson/Lab5_Poisson_Regression_Q.qmd)

## Instructions

-   If you are fitting a model, display the model output in a neatly formatted table. (The `tidy()` and `kable()` functions can help!)
-   If you are creating a plot, use clear labels for all axes, titles, etc.
-   Commit and push your work to GitHub regularly, at least after each exercise. Write short and informative commit messages.
-   When you're done, we should be able to knit the final version of the QMD in your GitHub as an HTML.

::: callout-tip
## Workflow tip

Work through the lab by running chunks **one at a time** in RStudio (Cmd/Ctrl + Enter), not by rendering the whole document. Render only at the end, once all your code is working.

This is how most R users actually work — you'd write a line, run it, check the output, and iterate. Rendering is for producing the final document. When you render, Quarto starts a fresh R session and runs every chunk top-to-bottom, so if anything is incomplete, everything downstream breaks. When you work interactively using chunks, your environment builds up as you go, so you always have your objects available to inspect and troubleshoot.

The **"Run All Chunks Above"** button (downward arrow with a bar, next to the green play button in each chunk) is another useful tool. It re-runs everything from the top of the document up to your current chunk, so all your objects are up to date before you start working on the next question.
:::

## Setup

```{r}
#| code-fold: false
library(tidyverse)
library(MASS)
library(naniar)
library(skimr)
library(emmeans)
library(ggeffects)
library(easystats)
library(performance)
library(knitr)
library(modelsummary)
library(pscl)

options(scipen = 999)
```

------------------------------------------------------------------------

# Part 1: Data Preparation & Cleaning

This lab accompanies a pre-registration document. Download it here: [Pre-Registration (PDF)](https://github.com/suyoghc/PSY504_Spring_2026/raw/main/posts/Poisson/Lab_Poisson_Pre-Registration.pdf). Refer to Hypothesis (a) for the variables and analysis plan you'll be implementing.

The data come from the 2010 General Social Survey (GSS). As described in the pre-registration, we are testing whether the number of hours per week a person spends on the Internet (`wwwhr`) is predicted by their vocabulary (`wordsum`), age (`age`), sex (`sex`), religiosity (`reliten`), political orientation (`polviews`), and how often they work from home (`wrkhome`).

```{r}
#| code-fold: false
data <- read_delim("https://raw.githubusercontent.com/jgeller112/psy504-advanced-stats/main/slides/Poisson/data/2010.csv")
```

The following code uses the `naniar` package to clean the data by replacing survey-coded missing values with `NA`.

```{r}
#| code-fold: false
data_pos <- data %>%
  dplyr::select(wwwhr, wordsum, age, sex, reliten, polviews, wrkhome) %>%
  replace_with_na(.,
    replace = list(
      wwwhr = c(-1, 998, 999),
      wordsum = c(-1, 99),
      reliten = c(0, 8, 9),
      polviews = c(0, 8, 9),
      wrkhome = c(0, 8, 9),
      age = c(0, 98, 99)
    ))
```

### Exercise 1.1 {.unnumbered}

Explain what the `replace_with_na` code is doing. Why are values like -1, 998, and 999 being replaced? What would happen to your model if you left these values in the data?

::: {.callout-tip collapse="true" title="Hint: GSS coding conventions"}
In many large-scale surveys, numeric codes like -1, 98/99, 998/999 don't represent real values. They're placeholders for "inapplicable," "don't know," or "no answer." Check the GSS codebook if you want to see what each code means for each variable.
:::

### Exercise 1.2 {.unnumbered}

Recode `sex` and `reliten` as factors with meaningful labels. Should `reliten` be treated as ordered or nominal? Justify your choice.

::: {.callout-tip collapse="true" title="Hint: Recoding"}
`sex` is coded 1 = Male, 2 = Female in the GSS. For `reliten`, check the unique values after cleaning. You'll need to decide whether to treat it as an ordered factor or collapse categories. Use `mutate()` with `factor()` or `case_when()`.
:::

```{r}

```

## Missingness

### Exercise 1.3 {.unnumbered}

Inspect the cleaned data for missing values. How many complete cases do you have across all analysis variables?

```{r}
#| code-fold: false
skimr::skim(data_pos)
```

------------------------------------------------------------------------

# Part 2: Fitting a Poisson Model

## Mini-tutorial: Fitting and reading a Poisson model

Before you fit your model on the GSS data, here is a worked example on a toy dataset so you can see the full workflow.

**Toy data:**

```{r}
#| code-fold: false
toy <- tibble(
  visits = c(0, 1, 3, 2, 0, 5, 1, 4, 7, 2),
  age = c(22, 25, 35, 30, 20, 55, 28, 45, 60, 33)
)
```

**Fit the model:**

```{r}
#| code-fold: false
toy_model <- glm(visits ~ age, data = toy, family = poisson(link = "log"))
```

**Read the output on the log scale.** These coefficients are on the log-count scale. The intercept is the log of the expected count when age = 0. The slope is the change in log-count for a one-year increase in age.

```{r}
#| code-fold: false
model_parameters(toy_model) %>% print_html()
```

**Exponentiate to get IRRs.** Now the intercept is the expected count when age = 0, and the slope is the incidence rate ratio — the multiplicative change in the expected count per one-year increase in age.

```{r}
#| code-fold: false
model_parameters(toy_model, exponentiate = TRUE) %>% print_html()
```

**Get a predicted count.** This is the model's predicted number of visits for a 40-year-old, back on the count scale.

```{r}
#| code-fold: false
predict(toy_model, newdata = tibble(age = 40), type = "response")
```

Now apply this workflow to the GSS data.

### Exercise 2.1 {.unnumbered}

Fit a Poisson regression model predicting `wwwhr` from the predictors listed in the pre-registration. Display the results in a formatted table.

::: {.callout-tip collapse="true" title="Hint: glm() syntax"}
The key argument is `family = poisson(link = "log")`. Use `model_parameters() %>% print_html()` or `tidy() %>% kable()` for a clean table.
:::

```{r}

```

### Exercise 2.2 {.unnumbered}

Pick one observation from your data. Using the coefficients from your fitted model, manually compute the predicted count for that observation. Show your work, then confirm it matches the output of `predict(model, type = "response")` for that row.

::: {.callout-tip collapse="true" title="Hint: From log-counts to counts"}
The model estimates log(λ). So the predicted log-count for observation *i* is β₀ + β₁x₁ + β₂x₂ + ... Exponentiate to get the predicted count: λ = exp(β₀ + β₁x₁ + ...). Use `coef(model)` to extract the coefficients.
:::

```{r}

```

------------------------------------------------------------------------

# Part 3: Model Diagnostics

### Exercise 3.1 {.unnumbered}

Use the `performance` package to check your model. For each diagnostic panel, state what it's testing and whether your model passes or fails.

::: {.callout-tip collapse="true" title="Hint: Which function?"}
`performance::check_model()` produces a panel of diagnostic plots. You can select specific checks with the `check` argument — e.g., `check = c("pp_check", "outliers", "overdispersion")`.
:::

```{r}

```

### Exercise 3.2 {.unnumbered}

Identify outliers. How many observations are flagged, and what makes them outlying: high residuals, high leverage, or both?

::: {.callout-tip collapse="true" title="Hint: Outlier detection"}
`performance::check_outliers()` returns a logical vector you can use to filter your data. Try `which()` or `sum()` to count the flagged cases.
:::

```{r}

```

### Exercise 3.3 {.unnumbered}

Refit the model without the outliers. Compare the two coefficient tables: which estimates changed substantively, and does the overall story change?

::: {.callout-tip collapse="true" title="Hint: Filtering"}
Store the outlier check as an object, then use it to index your data: `data_clean <- data_pos[!outlier_flag, ]`. Refit the same `glm()` call on the filtered data and compare with `model_parameters()`.
:::

```{r}

```

```{r}
#| eval: false
model_parameters(your_model) %>%
  print_html()
```

------------------------------------------------------------------------

# Part 4: Overdispersion & Alternative Models

### Exercise 4.1 {.unnumbered}

Check for overdispersion and report the dispersion ratio. What specific property of the Poisson distribution is violated when the ratio is much greater than 1?

::: {.callout-tip collapse="true" title="Hint: Checking overdispersion"}
`performance::check_overdispersion()` returns a dispersion ratio and a p-value. A ratio near 1 means the Poisson assumption holds.
:::

```{r}

```

### Exercise 4.2 {.unnumbered}

Given the overdispersion you found, name a model family that can handle it and explain in 1–2 sentences why it works. Then fit that model to the data and display the results.

::: {.callout-tip collapse="true" title="Hint: An extra parameter"}
The Poisson model has no separate parameter for variance. What if you had a model that added one? Check the `MASS` package.
:::

```{r}

```

------------------------------------------------------------------------

# Part 5: Model Comparison

### Exercise 5.1 {.unnumbered}

Compare your Poisson and negative binomial models using a likelihood ratio test and a comparison table. Interpret the test result and the AIC/BIC values. What do they tell you about relative model fit?

::: {.callout-tip collapse="true" title="Hint: Two ways to compare"}
`performance::test_likelihoodratio()` gives you a formal test. For a publication-style table, use `modelsummary(list("Poisson" = m1, "Neg. Binomial" = m2), exponentiate = TRUE, stars = TRUE)`. AIC/BIC are reported automatically.
:::

```{r}

```

------------------------------------------------------------------------

# Part 6: Zero Inflation

### Exercise 6.1 {.unnumbered}

Test for zero inflation in your chosen model and interpret the result. In your answer, define zero inflation and distinguish between the two types of zeros, giving a concrete example of each using the `wwwhr` variable.

::: {.callout-tip collapse="true" title="Hint: Why zero?"}
`performance::check_zeroinflation()` compares observed vs. predicted zero counts. Think about why someone might report 0 hours of internet use — there could be more than one reason.
:::

```{r}

```

## Mini-tutorial: Fitting a zero-inflated model

The `zeroinfl()` function from the `pscl` package fits a two-part model. The left side of `|` is the count model (Poisson or negative binomial). The right side models the probability of a structural zero using logistic regression.

```{r}
#| code-fold: false
# Extending the doctor visits example from Part 2
# Some people avoid doctors entirely (structural zeros)
# Others see doctors but had 0 visits that month (sampling zeros)
# Adding a binary predictor: whether the person has insurance

toy_zi_data <- tibble(
  visits   = c(0, 0, 0, 0, 1, 3, 0, 2, 0, 7, 0, 0, 4, 1, 0),
  age      = c(22, 25, 35, 19, 40, 55, 28, 45, 23, 60, 20, 30, 50, 33, 27),
  insured  = c(0, 0, 1, 0, 1, 1, 0, 1, 0, 1, 0, 0, 1, 1, 0)
)

# Count model: visits ~ age (how many visits, given you go at all)
# Zero model:  ~ insured (does having insurance affect whether you go at all?)
toy_zi_model <- zeroinfl(visits ~ age | insured, dist = "negbin", data = toy_zi_data)

summary(toy_zi_model)
```

The output has two sections. "Count model coefficients" are interpreted like a regular negative binomial (log-counts, exponentiate for IRRs). "Zero-inflation model coefficients" are interpreted like logistic regression (log-odds of being a structural zero). In this example, uninsured people may be more likely to avoid doctors entirely, which is a different process from how often insured people visit.

### Exercise 6.2 {.unnumbered}

Based on your zero-inflation test, should you fit a zero-inflated model? If yes, fit one and compare it to your negative binomial. If no, explain why the negative binomial is sufficient.

::: {.callout-tip collapse="true" title="Hint: Zero-inflated syntax"}
The mini-tutorial above shows the syntax pattern. The left side of `|` is your count model formula, the right side models structural zeros. Use `dist = "negbin"` if your count model is negative binomial.
:::

```{r}

```

------------------------------------------------------------------------

# Part 7: Visualization

### Exercise 7.1 {.unnumbered}

Visualize your final model's predictions on both the log-lambda and mean-count scales. For each plot, explain in 1–2 sentences what the scale shows and when you would report one vs. the other.

::: {.callout-tip collapse="true" title="Hint: Two scales"}
`ggemmeans()` or `ggpredict()` from the `ggeffects` package. Counts are back-transformed by default. Show both in the panel-tabset.
:::

::: panel-tabset
## Log Lambda

```{r}

```

## Mean Count

```{r}

```
:::

------------------------------------------------------------------------

# Part 8: Report Your Conclusions

### Exercise 8.1 {.unnumbered}

Write a results paragraph following the reporting checklist from the lecture slides. Include a justification for why you used your final model over the Poisson.

::: {.callout-tip collapse="true" title="Hint: Reporting checklist"}
From the lecture: state the hypothesis and model (with link function), report log-count coefficients and IRRs with 95% CIs, include a model fit measure (McFadden R²), mention overdispersion and zero-inflation test results, and note if you switched models and why.
:::
