---
title: "PSY 504 2026: Bayes Lab 1 — From Tables to MCMC"
format:
  html:
    code-fold: true
editor: visual
---

Consider the following setting: A factory produces coins, each with a different bias $\theta_i$. You don't know which coin you received. You have prior beliefs about which coins the factory is likely to produce — $p(\theta_i)$. You flip the coin $N$ times, observe $Y$ heads, and use that data to figure out which coin you probably got — $p(\theta_i \mid \text{data})$.

In Monday's lecture, you saw a setting like this play out with a table of probabilities:

-   each row was a different coin (a different $\theta_i$), showing what it could produce — $p(\text{data} \mid \theta_i)$ — with one column for each possible outcome (0 heads, 1 head, … $N$ heads)

-   you then observed $Y$ heads out of $N$ tosses

-   you fixed the data (the $\text{data} = Y$ column) and scored each coin on how well it explained your observation — that's the likelihood, $p(\text{data} = Y \mid \theta_i)$

-   then you multiplied each cell of this column by your prior on $\theta_i$, and divided by the column sum:

$$\frac{p(\text{data} = Y \mid \theta_i) \times p(\theta_i)}{\sum_i p(\text{data} = Y \mid \theta_i) \times p(\theta_i)}$$

-   this is the posterior — you just applied Bayes' rule!

In this lab, you'll do all of this in R — first by hand, then with `brms`. The entire lab uses coins. But you'll be using three different methods to obtain your posterior — a table, grid approximation, and MCMC.

***A note on how to work through this lab:*** Throughout this lab, you'll find that some code is commented out, and some is left for you to write yourself. You can uncomment the commented code to run it (CMD + Shift + C on selected lines). But before doing so, take a moment to think about how you'd write this yourself. What functions would you use? What inputs are needed, etc? Then uncomment the code, and make sure you understand each line / contrast with your approach before moving on. For the blanks marked \# Your code here: that's your turn — write the code, run it, and check that the output makes sense before continuing. Also, make sure to answer the descriptive follow-up questions such as "what does these tell you?"\
\

## Setup

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(ggdist)
```

# Part 1: The lecture table — in R

In lecture, we built a table with 5 coins ($\theta_i$ = 0.2, 0.4, 0.5, 0.6, 0.8) and $N = 5$ flips, giving 6 possible outcomes (0 through 5 heads). Let's recreate it, and then observe $Y = 3$ heads.

## Section 1: Build the probability table

Use `dbinom()` to compute the probability of each outcome for each coin. (This is reading **across a row** — $p(\text{data} \mid \theta_i)$ with $\theta_i$ fixed.)

```{r}
# ## Define our coins
# theta_values <- c(0.2, 0.4, 0.5, 0.6, 0.8)
# 
# ## For each theta_i, compute the probability of 0, 1, 2, 3, 4, 5 heads
# prob_table <- tibble(theta = theta_values) %>%
#   rowwise() %>%
#   mutate(
#     probs = list(dbinom(0:5, size = 5, prob = theta))
#   ) %>%
#   unnest_wider(probs, names_sep = "_") %>%
#   rename_with(~ paste0("heads_", 0:5), starts_with("probs_"))
# 
# prob_table
```

### Question 1.1

Verify that each row sums to 1. What does this tell you?

```{r}
# Your code here: using a similar structure as above, compute row sums, mutate it (and excluding the theta column)
```

## Section 2: Read the table down a column

Now fix the data at 3 heads (Y = 3) and read **down the column**. This gives you the likelihood function for Y = 3 — $p(\text{data} = 3 \mid \theta_i)$ — which scores each coin on how well it explains your observation.

Hint: you used `dbinom()` in Section 1 across outcomes — now use it across $\theta_i$ values, fixing the number of heads at 3.

::: {.callout-tip collapse="true" title="Code"}
```{r}
# ## Score each coin on how well they explain 3 heads out of 5 (Y = 3, N = 5)
# likelihood_table <- tibble(
#   theta = theta_values,
#   likelihood = dbinom(3, size = 5, prob = theta_values)
# )
# 
# likelihood_table
```
:::

### Question 2.1

Does the likelihood column sum to 1? Compute it. What does this tell you about the likelihood?

```{r}
# Your code here
```

### Question 2.2

Which $\theta_i$ has the highest likelihood score? That's the MLE. Can you call it a probability? Why or why not?

## Section 3: Compute the posterior with a flat prior

Now do what Bayes' rule does: multiply each likelihood by a prior, then normalize so it sums to 1.

Start with a flat prior — each $\theta_i$ is equally plausible before seeing data. You have the `likelihood_table` from Section 2. Add columns for the prior, the product of prior × likelihood, and the normalized posterior.

::: {.callout-tip collapse="true" title="Code"}
```{r}
# posterior_table <- likelihood_table %>%
#   mutate(
#     prior = 1 / n(),                          # flat prior: each theta_i gets equal weight
#     prior_x_likelihood = prior * likelihood,   # numerator of Bayes' rule
#     posterior = prior_x_likelihood / sum(prior_x_likelihood)  # normalize
#   )
# 
# posterior_table
```
:::

### Question 3.1

Verify that the posterior column sums to 1. What does this tell you?

```{r}
# Your code here
```

### Question 3.2

In lecture, we asked: "There's a \_\_\_% chance that $\theta$ is 0.6." Fill in the blank using your posterior table. Can you also state: "There's a \_\_\_% chance that $\theta$ is 0.5 or 0.6"?

```{r}
# Your code here
```

You just did something that the likelihood alone could not do — you made a probability statement about a parameter.

## Section 4: Change the prior

Now use an informative prior — "most coins are roughly fair." Assume that the bias probabilities are c(0.05, 0.20, 0.50, 0.20, 0.05)

```{r}
# posterior_informative <- likelihood_table %>%
#   mutate(
#     # Prior peaked at 0.5, lower at extremes
#     prior = c(0.05, 0.20, 0.50, 0.20, 0.05),
#     prior_x_likelihood = prior * likelihood,
#     posterior = prior_x_likelihood / sum(prior_x_likelihood)
#   )
# 
# posterior_informative
```

### Question 4.1

Compare this posterior to the flat-prior posterior. Which $\theta$ is most probable now? Did the prior pull the estimate? In which direction?

### 4.2

Plot both posteriors side by side.

```{r}
# Combine the two posteriors for plotting
bind_rows(
  posterior_table %>% mutate(prior_type = "Flat prior"),
  posterior_informative %>% mutate(prior_type = "Informative prior")
) %>%
  ggplot(aes(x = factor(theta), y = posterior, fill = prior_type)) +
  geom_col(position = "dodge") +
  labs(x = expression(theta), y = "Posterior probability",
       fill = "Prior") +
  theme_minimal()
```

### Question 4.3

The data are the same (3 heads out of 5). Only the prior changed. Why did the posterior shift? When would this matter more — with 5 flips or 500 flips?

# Part 2: Grid approximation — finer resolution

With only 5 values of $\theta_i$, we got a rough picture. What if we tried 1,000 values between 0 and 1? That's grid approximation — the same logic, just with a much finer grid. Now $\theta$ becomes effectively continuous.

## Section 5: Grid approximation with a flat prior

Replace the 5 $\theta_i$ values with `seq(0, 1, length.out = 1000)` and repeat the same multiply-and-normalize steps from Section 3. Then plot the posterior.

::: {.callout-tip collapse="true" title="Code"}
```{r}
# ## Define a fine grid of theta values
# grid <- tibble(
#   theta = seq(0, 1, length.out = 1000)
# ) %>%
#   mutate(
#     prior = 1,                                           # flat prior (unnormalized is fine)
#     likelihood = dbinom(3, size = 5, prob = theta),      # score each theta
#     unstd_posterior = prior * likelihood,                 # prior × likelihood
#     posterior = unstd_posterior / sum(unstd_posterior)     # normalize
#   )
# 
# ## Plot the posterior
# ggplot(grid, aes(x = theta, y = posterior)) +
#   geom_line() +
#   labs(x = expression(theta), y = "Posterior probability",
#        title = "Posterior: 3 heads out of 5 (Y = 3, N = 5), flat prior") +
#   theme_minimal()
```
:::

### Question 5.1

Where is the peak of the posterior? Does this match the MLE from Section 2?

## Section 6: Grid approximation with an informative prior

Now replace the flat prior with a Beta(10, 10) distribution — peaked near 0.5, skeptical of extreme values.\
\
***Note below:** The likelihood and posterior are on different scales, so we rescale the likelihood to the same height for visual comparison. It's juat a plotting/visualization trick. The shapes and peaks are what matter here, not the y-axis values.*

```{r}
grid_informative <- tibble(
  theta = seq(0, 1, length.out = 1000)
) %>%
  mutate(
    prior = dbeta(theta, 10, 10),                        # peaked near 0.5
    likelihood = dbinom(3, size = 5, prob = theta),
    unstd_posterior = prior * likelihood,
    posterior = unstd_posterior / sum(unstd_posterior)
  )

## Plot prior, likelihood (rescaled), and posterior together
grid_informative %>%
  mutate(
    likelihood_scaled = likelihood / max(likelihood) * max(posterior)  # rescale for visibility
  ) %>%
  ggplot(aes(x = theta)) +
  geom_line(aes(y = posterior), color = "blue", linewidth = 1) +
  geom_line(aes(y = prior / sum(prior)), color = "red", linetype = "dashed") +
  geom_line(aes(y = likelihood_scaled), color = "gray50", linetype = "dotted") +
  labs(x = expression(theta), y = "Density (rescaled)",
       title = "Blue = Posterior, Red = Prior, Gray = Likelihood") +
  theme_minimal()
```

### Question 6.1

The posterior (blue) sits between the prior (red) and the likelihood (gray). Why? What would happen to the posterior if you had $Y = 300$ heads out of $N = 500$ instead of $Y = 3$ out of $N = 5$?

## Section 7: Watch the prior wash out

Increase the sample size while keeping the same proportion of heads (60%).

```{r}
# Same proportion (60% heads), increasing sample sizes
sample_sizes <- c(5, 20, 50, 200)
heads_observed <- round(0.6 * sample_sizes)  # Y = 3, 12, 30, 120

grid_by_n <- map2_dfr(sample_sizes, heads_observed, function(n, k) {
  tibble(
    theta = seq(0, 1, length.out = 1000),
    n = n,
    prior = dbeta(theta, 10, 10),
    likelihood = dbinom(k, size = n, prob = theta),
    unstd_posterior = prior * likelihood,
    posterior = unstd_posterior / sum(unstd_posterior)
  )
})

ggplot(grid_by_n, aes(x = theta, y = posterior)) +
  geom_line(color = "blue") +
  geom_line(aes(y = prior / sum(prior)), color = "red", linetype = "dashed",
            data = grid_by_n %>% filter(n == 5)) +
  facet_wrap(~ paste0("n = ", n), scales = "free_y") +
  labs(x = expression(theta), y = "Posterior probability",
       title = "Same prior, more data → posterior concentrates") +
  theme_minimal()
```

### Question 7.1

What happens to the posterior as n increases? Does the prior still matter at n = 200? This is what the "Three ingredients" slide was about: "With lots of data — likelihood dominates."

# Part 3: Let brms do it

Grid approximation works beautifully for one parameter. But your regression models have $\beta_0$, $\beta_1$, and $\sigma$ — three parameters. A 3D grid with 1000 points per dimension means \_\_\_\_\_\_\_(fill in the blank) cells. That's why we need MCMC. Let's fit the same coin problem with `brms` and see that it gives us the same answer.

## Section 8: Set up the coin data

`brms` needs data in a data frame. For a binomial model, we tell it how many successes and how many trials.

```{r}
# Our coin data: Y = 3 heads out of N = 5 flips
coin_data <- data.frame(
  heads = 3,
  trials = 5
)
```

## Section 9: Fit with brms — flat prior

We'll use `family = binomial` and a very wide prior on the intercept (which is on the logit scale). A wide prior on the logit scale approximates a flat prior on the probability scale.

```{r fit_flat, results = "hide", message = FALSE}
fit_flat <- brm(
  heads | trials(trials) ~ 1,     # intercept-only binomial model
  family = binomial(link = "logit"),
  data = coin_data,
  prior = prior(normal(0, 10), class = Intercept),  # wide prior ≈ flat on prob scale
  seed = 504,
  iter = 4000, warmup = 1000, chains = 4
)
```

```{r}
summary(fit_flat)
```

### Question 9.1

The intercept is on the logit scale. To convert to $\theta$ (probability scale), apply the inverse logit: `plogis()`. What's the posterior mean of $\theta$? Is it closer to 0.5 or 0.6? Why?

```{r}
# # Extract posterior draws of the intercept
# draws_flat <- as_draws_df(fit_flat) %>%
#   mutate(theta = plogis(b_Intercept))  # convert logit → probability
# 
# # Posterior mean of theta
# mean(draws_flat$theta)
```

### Question 9.2

How does this compare to your grid approximation result from Section 5?

## Section 10: Plot the posterior from MCMC draws

```{r}
# ggplot(draws_flat, aes(x = theta)) +
#   geom_histogram(aes(y = after_stat(density)), bins = 50,
#                  fill = "steelblue", alpha = 0.7) +
#   labs(x = expression(theta), y = "Density",
#        title = "MCMC posterior: 3 heads out of 5 (Y = 3, N = 5), approximately flat prior") +
#   theme_minimal()
```

### Question 10.1

This histogram is made from 12,000 draws (4 chains × 3000 post-warmup). Each draw is a value of $\theta$ sampled from the posterior. The histogram IS the posterior. Compare the shape to your grid approximation plot from Section 5 — how does the shape compare to your grid approximation plot from Section 5? What's similar? What's different?

## Section 11: Fit with brms — informative prior

Now use a prior that says "most coins are roughly fair." On the logit scale, $\theta = 0.5$ corresponds to logit = 0. A Normal(0, 0.5) prior on the logit scale concentrates $\theta$ near 0.5.

```{r fit_informative, results = "hide", message = FALSE}
fit_informative <- brm(
  heads | trials(trials) ~ 1,
  family = binomial(link = "logit"),
  data = coin_data,
  prior = prior(normal(0, 0.5), class = Intercept),  # informative: most coins near fair
  seed = 504,
  iter = 4000, warmup = 1000, chains = 4
)
```

```{r}
draws_informative <- as_draws_df(fit_informative) %>%
  mutate(theta = plogis(b_Intercept))

mean(draws_informative$theta)
```

### Question 11.1

The posterior mean shifted compared to the flat prior. In which direction? Why? Does this match what happened in Section 4 when you changed the prior by hand?

## Section 12: Overlay both posteriors

```{r}
bind_rows(
  draws_flat %>% select(theta) %>% mutate(prior_type = "Wide (≈ flat)"),
  draws_informative %>% select(theta) %>% mutate(prior_type = "Informative")
) %>%
  ggplot(aes(x = theta, fill = prior_type)) +
  geom_density(alpha = 0.5) +
  labs(x = expression(theta), y = "Density",
       fill = "Prior",
       title = "Same data, different priors → different posteriors") +
  theme_minimal()
```

### Question 12.1

With only 5 flips, the prior matters. What would you expect if you had 500 flips? (You can test this: change the data to `heads = 300, trials = 500` and refit both models.)

## Section 13: Fill in the blank

Using your posterior draws, make the probability statements that MLE cannot. Hint: what fraction of your draws satisfy a given condition? Think about what `mean()` returns when applied to a logical vector.

::: {.callout-tip collapse="true" title="Code"}
```{r}
# "What's the probability that theta > 0.5?"
mean(draws_flat$theta > 0.5)

# "What's the probability that theta is between 0.4 and 0.8?"
mean(draws_flat$theta > 0.4 & draws_flat$theta < 0.8)
```
:::

### Question 13.1

In lecture, the blank was: "There's a \_\_\_% chance that $\theta$ is 0.6." With continuous posterior draws, we can't ask about a single point. Instead, we ask about intervals. Compute: "There's a \_\_\_% chance that $\theta$ is between 0.55 and 0.65." How does this compare to the grid approximation posterior at $\theta = 0.6$?

## Section 14: Credible intervals

The 95% credible interval is the central 95% of the posterior draws. This has the interpretation that frequentist confidence intervals don't: "There's a 95% probability that $\theta$ lies within this interval."

```{r}
# # 95% credible interval
# quantile(draws_flat$theta, probs = c(0.025, 0.975))
```

### Question 14.1

Is $\theta = 0.5$ inside the 95% credible interval? What about $\theta = 0.2$? What does this tell you?

# Wrapping up

You just solved the same problem three ways:

1.  **By hand** (5 $\theta_i$ values, a table, multiply and normalize)
2.  **Grid approximation** (1,000 $\theta$ values, same logic, finer resolution)
3.  **MCMC via brms** (12,000 draws from the posterior, no grid needed)

All three gave you the same answer. The first two don't scale to real models with many parameters. The third does. That's why we use `brms`.

### Final reflection

In your own words, explain the connection between the posterior table from Section 3 and the histogram from Section 10. What's the same? What's different?

## References

Kurz, A. S. (2025). *Statistical rethinking with brms, ggplot2, and the tidyverse: Second edition* (Version 1.2.0). <https://bookdown.org/content/4857/>

McElreath, R. (2020). *Statistical rethinking: A Bayesian course with examples in R and Stan* (Second Edition). CRC Press.

## Session information

```{r}
sessionInfo()
```
